Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-08-30 write-up/covid-writeup.tex
Update covid-writeup.tex
closer- just getting question right

In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.  Weight estimation requires auxiliary information provided by random samples.  A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in data quantity.

@@

When an individual is infected with SARS-CoV-2, there will typically be an incubation period which may last between two to five days (cite: NPR) before the viral load is high enough to be detected.  A \emph{molecular test} refers to tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).  If someone has an active infection, then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).  After that, the viral load will decrease and a molecular test with perfect specificity will come back negative.

While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease after the first few days after infection, meaning a positive molecular test does not imply transmissibility. This is of particular importance for epidemiological forecasts built from population-level data and impacts our model choice in Section~\ref{section:modelbased}.

Molecular tests are referred to as diagnostic tests.

Serological (or antibody) tests look for an immunological response to the virus.  A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.  At this point, a serological test with perfect sensitivity will come back positive.  This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.

Seroprevalence studies are useful for disease surveillance.

To date, however, the number of seroprevalance surveys are limited.

See XX for further discussion.

Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent further spread.  ,

Significant logistic and capacity constraints have

Our primary goal is to estimate accurately estimate the fraction of active infections in the population.

This is the most relevant
Secondary goals include estimation of rates of change and the effective reproduction number which are also useful in policy making.

Alternatively,
Such testing would result in higher detection rates.
Will these results be reported to local authorities.
If not, selection bias in who uses these tests and then reports the results is likely to matter.

When an individual is infected with SARS-CoV-2 there is a

A viral test such as the RT-PCR

Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a

\subsection{COVID-19 case counts, viral testing counts, and deaths}
\label{section:testinginfo}

Coronavirus case-count data refers to the number of positive diagnostic tests performed on each day.  Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Departments of Health in each state also typically report the total number of reported tests per day.  On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide incoming county-level information of confirmed cases and deaths.

{\bf key point: WHAT IS THE GOAL OF CASE COUNT DATA? WHAT IS THE GOAL OF SEROPREVALENCE STUDIES?  are we using it for that reason. cse count data is colelcted as part of surveillance programs wehre teh goal is to identify and isolate active infections.  However, many people are trying to use these to udnerstand the disease.  This is our issue.}

@@

>>>>

In Section~\ref{section:improvedcasecount}, we present a doubly robust estimation method for combining epidemiological forecasts with inverse-probability weighting methods to estimate prevalence.  Weight estimation requires auxiliary information provided by random samples.  A statistical error decomposition illustrates a trade-off in use of weights: a potential increase in data quality but a reduction in effective data quantity.

@@

After an individual is infected with SARS-CoV-2, there is an incubation period which lasts between two to five days (cite: NPR) before the viral load is high enough to be detected.  A \emph{molecular test} refers to diagnostic tests that aim to detect increased viral load (e.g., RT-PCR or antigen tests).  If someone has an active infection then after the incubation period a molecular test with perfect sensitivity will yield a positive result for the next few weeks (cite: wlos).  After that, the viral load will decrease and a molecular test with perfect specificity will come back negative. While an infected individual may yield a positive molecular test for several weeks, they will likely stop transmitting the disease within a few days of infection, meaning a positive molecular test does not imply transmissibility.

A molecular test conducted on an infected individual may return a negative result.  Such false negatives are very strongly associated with when the test is conducted.  In the incubation phase, most molecular tests will return a false negative result.  Molecular tests are most sensitive when the viral loads are highest which occurs during the first few days of transmissibility (cite: ).   Moreover, most molecular tests are performed via nasopharyngeal swab (cite: NEJM).  Specimen collection by swab is known to impact false negative/positive rates regardless of test timing.   Systematic reviews suggest that 70\% sensitivity and 95\% specificity are reasonable estimates for current RT-PCR tests (cite: NEJM).

The primary goal is of molecular tests is in diagnosis of active infections in the population.  Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation. Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent systemic outbreaks.  There has been work on developing faster, cheaper testing options (cite: Mina) that can be used at-home as well as group testing (citE: Mina); however, these developments are beyond the scope of this paper.  Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.

\emph{Serological tests} look for an immunological response to the virus.  A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.  At this point, a serological test with perfect sensitivity will come back positive.  This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.

To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.  These studies are useful for disease surveillance. When the number of active infections in a population are low, serological surveys are cheaper and more accurate than molecular tests and are more useful in detecting outbreaks (cite: Mina). To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.  While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.  Here we consider whether non-probability diagnostic testing can be used in estimation of prevalence of immunity.

\subsection{Publicly available data on COVID-19}
\label{section:testinginfo}

In this paper, our primary goal is to accurately estimate the active infection rate in the population over time using \emph{publicly available} test results.  Secondary goals include estimation of rates of change in these quantities and the effective reproduction number (cite:) which characterize the infectious disease spread.  A final goal is using these estimates to characterize prevalence of immunity in the general population.  As previously stated, these quantities are fundamental to public health policy and general critical information on presence and transmission of SARS-CoV-2 as well as the impact of public health  interventions.

% {\bf key point: WHAT IS THE GOAL OF CASE COUNT DATA? WHAT IS THE GOAL OF SEROPREVALENCE STUDIES?  are we using it for that reason. cse count data is colelcted as part of surveillance programs wehre teh goal is to identify and isolate active infections.  However, many people are trying to use these to udnerstand the disease.  This is our issue.}

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~\ref{fig:ny-covid}, dotted blue line). On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

@@

This is of particular importance for epidemiological forecasts built from population-level data and impacts our model choice in Section~\ref{section:modelbased}.
{\bf Check if SIR models account for this?}
